---
title: 'Case Study 02'
author: 'André Victor Ribeiro Amaral'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction

For this problem, we will study how the amount of carbon fiber and sand additions affect the hardness of a certain material. You can download the `.csv` file [here](./datasets/hardness.csv).

This data set contains following the variables:

1. `sand` (multi-valued discrete)
2. `carbon` (multi-valued discrete)
2. `hardness` (continuous)

# Exploring the data set

In order to explore the data set and perform initial analyses, we have to read it (with `R`) first. Provided that the `.csv` file is saved within the `datasets/` folder, one can read the file in the following way.

```{r}
col.names <- c('sand', 'carbon', 'hardness')
hardness <- read.csv(file = 'datasets/hardness.csv', header = FALSE, sep = ',', col.names = col.names)
head(hardness, 5)
```

Now, let's see a summary of our data.

```{r}
summary(hardness)
```

As one can see from the above table, the multi-valued discrete attributes are being interpreted as numeric; however, we should set them as factor. For changing them, we can do the following

```{r}
hardness$sand <- factor(hardness$sand, levels = c(0, 15, 30), labels = c('lowS', 'medS', 'highS'))
hardness$carbon <- factor(hardness$carbon, levels = c(0.00, 0.25, 0.50), labels = c('lowC', 'medC', 'highC'))
summary(hardness)
```

Once we have the data, we can do a few plots that might help us understanding our problem. We can start by doing two different boxplots for the two categorical variables.

```{r}
boxplot(hardness ~ sand, data = hardness)
boxplot(hardness ~ carbon, data = hardness)
```

From both plots, it seems like there is a difference in one of the treatments (low doses for both sand and carbon) regarding the hardness of the material. However, we should only analyse the two factors separately if we can conclude that there is no interaction between them, meaning that there is no partial effect of one factor over the other. To see this, let's use the `interaction.plot()` function to explore the possible interactions between the factors.

```{r}
interaction.plot(x.factor = hardness$sand, trace.factor = hardness$carbon, response = hardness$hardness)
interaction.plot(x.factor = hardness$carbon, trace.factor = hardness$sand, response = hardness$hardness)
```

From the both plots above, the interactions (if any) do not seem to be important (since the lines are, more or less, parallel). As an important side note, if there is **no** interaction, the main effects (if significant) will tell us if there is variation among the levels of the factor in question; however, if modeled with interaction, the main effects may be difficult to be correctly explored and should be avoided. In this case, one may simply consider all pairs of levels as separate treatments.


# Fitting a model

So let's start by fitting our complete model.

```{r}
model <- lm(formula = hardness ~ sand * carbon, data = hardness)
anova(model)
```

As we can see from the above table (and p-values), we **reject** the hypothesis "sand have no significant effect on hardness", we reject the hypothesis "carbon have no significant effect on hardness", as we **fail to reject** the hypothesis "sand and carbon interaction have no significant on hardness". Thus, we can fit the following simplified model

```{r}
model2 <- lm(formula = hardness ~ sand + carbon, data = hardness)
anova(model2)
```

Now, as [before](./case_study_01.html), we can compute the estimated means and effects for sand and carbon from `hardness ~ sand + carbon`.

```{r}
modAOV <- aov(formula = hardness ~ sand + carbon, data = hardness)
model.tables(modAOV, 'means')

# Effects
model.tables(modAOV)
```

Next, let's do a residual analysis, so that we can make sure that our assumptions hold. We will plot the "Residuals vs. Fitted" and "Normal Q-Q" graphs, and conduct the appropriate tests, namely, `ncvTest()` (notice it is not the same as `leveneTest()`) and `shapiro.test()` (for `resid()`), for the equal variance and normally distributed residuals assumptions, respectively.

```{r}
plot(model2, which = c(1, 2))

library('car')
ncvTest(model2)
shapiro.test(resid(model2))
```

As the p-values are large for both tests, the assumptions seem to hold. Also, the plots allow us to conclude the same. Therefore, `model2` might correctly model our problem.

Also, as [before](./case_study_01.html), we can also do a pairwise analysis.

```{r}
(modTuk <- TukeyHSD(modAOV))
plot(modTuk)
```

# Predicting unknown values

Finally, the predicted values for each combination of level of `sand` and `carbon` can be computed using the `predict()` function.

```{r}
(newdata <- data.frame(sand = factor(rep(c('lowS', 'medS', 'highS'), each = 3)), carbon = factor(rep(c('lowC', 'medC', 'highC'), times = 3))))
newdata$predict <- predict(model2, newdata = newdata)
newdata
```
