---
title: 'Case Study 03'
author: 'André Victor Ribeiro Amaral'
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

# Introduction

For this problem, we will analyse data collected in an observational study in a semiconductor manufacturing plant. Data were retrieved from the [**Applied Statistics and Probability for Engineers** book](https://www.wiley.com/en-us/Applied+Statistics+and+Probability+for+Engineers%2C+7th+Edition-p-9781119400363). You can download the `.csv` file [here](./datasets/wire_bond.csv). In this plant, the finished semiconductor is wire-bonded to a frame. The variables reported are pull strength (a measure of the amount of force required to break the bond), the wire length, and the height of the die. Thus, the data set contains following the attributes:

1. `pull_strength` (quantitative)
2. `wire_length` (quantitative)
2. `height` (quantitative)

# Exploring the data set

In order to explore the data set and perform initial analyses, we have to read it (with `R`) first. Provided that the `.csv` file is saved within the `datasets/` folder, one can read the file in the following way.

```{r}
col.names <- c('pull_strength', 'wire_length', 'height')
data <- read.csv(file = 'datasets/wire_bond.csv', header = FALSE, sep = ',', col.names = col.names)
head(data, 5)
```

Now, let's see a summary of our data.

```{r}
summary(data)
```

In this case, there are no missing values---which is great. However, since we want to use most information from this data set, it might not be easy to visualize how one attribute (in our case `pull_strength`) can be written as a function of more than two variables at the same time. Fortunately, we can still plot data in 3D (using the `scatterplot3d()` function from the `scatterplot3d` package).

```{r}
library('scatterplot3d')
scatterplot3d(data$wire_length, data$height, data$pull_strength, color = 2, pch = 19, angle = 70)
```

Also, it is useful to see how variables are correlated. To do this, we can plot (at least) two types of graphs; in this case, we will use the `scatterplotMatrix()` function from the `car` package, and the `corrplot.mixed()` function from the `corrplot` package to plot the correlation (from `cor()`) matrix.

```{r}
library('car')

scatterplotMatrix(data)
```

```{r}
library('corrplot')

corrplot.mixed(cor(data))
```

From the second plot, we can see that `pull_strength` is highly correlated with `wire_length`, and this relation is detailed in the first plot. These graphs will be even more useful once we have more variables (as we will see in our next problem).

# Fitting a model

Our very first task will be fitting a model with all variables so that we can try to explain how `pull_strength` relates with the other variables. We can do this in the following way.

```{r}
model <- lm(formula = pull_strength ~ ., data = data)
summary(model)
```

As we can see from the summary, for a significance level of $5\%$, all coefficients are different than zero. Thus, we should keep them (for our next problem, we will try to simplify the model further).

Once we have the model, a next step would be performing a residual analysis. To do this, we can plot the diagnostic graphs and run the appropriate tests.

```{r}
plot(model, which = c(1, 2))
```

The plot seems okay, but we still have to do the tests. As a remark, and as extracted from [this page](https://stats.stackexchange.com/questions/239157/what-does-the-residuals-vs-fitted-regression-line-express), "*the red line is a LOWESS fit to your residuals vs fitted plot. Basically, it's smoothing over the points to look for certain kinds of patterns in the residuals. For example, if you fit a linear regression on data that looked like $y = x^2$, you'd see a noticeable bowed shape*". Regarding the tests, we have the following

```{r}
ncvTest(model)
shapiro.test(resid(model))
```

Also, from the test results, we fail to reject the equal variance and normality assumptions, meaning that we have a good model for our data.

In particular, the model is given by

\begin{align*}
\texttt{pull_strength}_i &= `r round(model$coefficients[[1]], 3)` + `r round(model$coefficients[[2]], 3)`\texttt{wire_length}_i + `r round(model$coefficients[[3]], 3)`\texttt{height}_i
\end{align*}

And we can plot the fitted model in the following way

```{r}
plot3d <- scatterplot3d(data$wire_length, data$height, data$pull_strength, color = 2, pch = 19, angle = 70)
plot3d$plane3d(model)
```


# Predicting unknown values

Now that we have a "good" fitted model, we can predict, the value of `pull_strength` for new values of `wire_length` and `height`. For instance, we can predict the value of `pull_strength`, such that `wire_length` is equal to 7.5 and 12.5 and `height` is equal to 150 and 450, respectively. **Notice that your new points must lie within the range for the observed data**, otherwise your model may not be appropriate for predicting extrapolated points.

```{r}
newdata <- data.frame(wire_length = c(7.5, 12.5), height = c(150, 450))

(pred1 <- predict(object = model, newdata = newdata, interval = 'confidence'))
(pred2 <- predict(object = model, newdata = newdata, interval = 'prediction'))

newdata <- cbind(newdata, pull_strength = pred1[, 1])
data <- rbind(data, newdata)

plot3d <- scatterplot3d(data$wire_length, data$height, data$pull_strength, color = c(rep(2, nrow(data) - 2), 3, 3), pch = 19, angle = 70)
plot3d$plane3d(model)
legend('topleft', c('Observed data', 'New data'), col = c(2, 3), pch = 19)
```

From the above predicted values, notice that this time, we added the confidence interval and the prediction interval. As explained [Applied Statistics and Probability for Engineers book](https://www.wiley.com/en-us/Applied+Statistics+and+Probability+for+Engineers%2C+7th+Edition-p-9781119400363), "*the prediction interval is always wider than the confidence interval. The confidence interval expresses the error in estimating the mean of a distribution, and the prediction interval expresses the error in predicting a future observation from the distribution at the point $\mathbf{x}_0$. This must include the error in estimating the mean at that point as well as the inherent variability in the random variable `wire_length` at the same value $\mathbf{x} = \mathbf{x}_0$*".